---
title: "Bayes Expression Variance"
author: "Julin N Maloof"
date: "6/21/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
library(brms)
library(edgeR)
library(tidyverse)
```

Want to fit a model where:

* each gene has an intercept
* each gene has its own variance

## preliminaries

### get the data

You can also download it [here](http://jnmaloof.github.io/BIS180L_web/data/gh_internode_counts.tsv)

```{r}
counts.data <- read.delim("~/git/BIS180L_web/data/gh_internode_counts.tsv")
```

### create a data frame that describes the samples

```{r}
sample.description <- data.frame(
  sample=colnames(counts.data),
  
  #This next line searches for IMB211 or R500 in the colnames of counts.data and returns anything that matches
  #In this way we can extract the genotype info.
  gt=regmatches(colnames(counts.data),regexpr("R500|IMB211",colnames(counts.data))),
  
  #Now we use the same method to get the treatment
  trt=regmatches(colnames(counts.data),regexpr("NDP|DP",colnames(counts.data)))
)

# Now we can paste the trt and gt columns together to give a group identifier
sample.description$group <- paste(sample.description$gt,sample.description$trt,sep="_")

# set the reference treatment to "NDP"
sample.description$trt <- relevel(sample.description$trt,ref="NDP")

sample.description
```

### normalize counts 

```{r}
counts.data <- counts.data[rownames(counts.data)!="*",]
counts.data[is.na(counts.data)] <- 0
counts.data <- counts.data[rowSums(counts.data > 10) >= 3,]
dge.data <- DGEList(counts=counts.data, group=sample.description$group)
dim(dge.data) 
dge.data <- calcNormFactors(dge.data, method = "TMM")
dge.data$samples # look at the normalization factors
```

```{r}
data.cpm <- as.tibble(cbind(geneID=rownames(cpm(dge.data)),as.tibble(cpm(dge.data)),stringsAsFactors=FALSE))
colnames(data.cpm) <- sub("\\.1.*","",colnames(data.cpm))
head(data.cpm)
```

### rearrange cpm

```{r}
data.tidy <- data.cpm %>% 
  gather(sampleID,cpm,-geneID) %>% 
  separate(sampleID,into=c("gt","trt","rep","tissue"))
head(data.tidy)
```

## get started

Start with voom transformed data and Gaussian distribution just to get our head around this
```{r}
voom.data <- voom(dge.data)$E
voom.data <- as.tibble(cbind(geneID=rownames(voom.data),as.tibble(voom.data),stringsAsFactors=FALSE))
colnames(voom.data) <- sub("\\.1.*","",colnames(voom.data))
voom.tidy <- voom.data %>% 
  gather(sampleID,expression,-geneID) %>% 
  separate(sampleID,into=c("gt","trt","rep","tissue"))
head(voom.tidy)
voom.small <- voom.tidy %>% filter(gt=="IMB211" & trt=="DP" & startsWith(geneID,"Bra0000"))
hist(voom.small$expression)
dim(voom.small)
```

What priors can we use in standard Gaussian?
```{r}
get_prior(expression ~ (1|geneID), data=voom.tidy)
```

### Standard Gaussian model with gene as random effect and non-hierarchical prior for sigma

```{r, results="hide"}
system.time(fit1 <- brm(expression ~ (1|geneID), 
            prior=c(
              set_prior("normal(4,1)","Intercept"),
              set_prior("student_t(3,0,1)","sd"),
              set_prior("student_t(3,0,1)","sigma")
              ),
            data = voom.small,
            iter=10000))
```

```{r}
fit1
plot(fit1)
```

### Gaussian, assigning different sigmas to different genes (As random effect)

what are the priors that can be set?
```{r}
get_prior(bf(expression ~ (1|geneID), sigma ~ (1|geneID)), data=voom.small)
```

Do the fit specifying geneID as a (random) predictor of sigma

Must increase adapt_delta and number of iterations to get good sampling.
```{r, resluts="hide"}
system.time(fit2 <- brm(bf(expression ~ (1|geneID), sigma ~ (1|geneID)),
            prior=c(
              set_prior("normal(4,1)","Intercept"),
              set_prior("student_t(3,0,1)","sd")),
            data = voom.small,
            iter=10000,
            warmup = 2000,
            control=list(adapt_delta=0.95)
))

```

```{r}
fit2
plot(fit2)
```

### Compare models with and without hiearchical prior for sigma:

```{r}
loo(fit1,fit2)
```
lower LOOIC is better, so there is strong evidence that having a hierarchical prior on sigma provides a better model fit.

### explore the fitted parameters

Get the estimated sigmas for each gene:

```{r}
str(ranef(fit2)) #we want the second one
# also, to get back on the original scale we need to take the exponential: see vignette("brms_distreg")
sigma.gene <- ranef(fit2)[[2]]
hist(sigma.gene) #how much the log_sigma is being adjusted for each gene
hist(exp(sigma.gene)) #in sd units
hist(exp(sigma.gene+fixef(fit2)["sigma_Intercept",])) #actual sigma being used for each gene
```

Get the estimated mean expression for each gene
```{r}
mean.gene <- ranef(fit2)[[1]] + fixef(fit2)["Intercept",]
```

Is there a relationship between expression level and standard deviation?  We hope that voom took care of this:

```{r}
mean.var <- data.frame(mean=mean.gene,sigma=exp(sigma.gene+fixef(fit2)["sigma_Intercept",]))
colnames(mean.var) <- c("mean","sigma")
ggplot(mean.var,aes(x=mean,y=sigma)) + geom_point() + geom_smooth()
```
Seems reasonable; no strong trend

To test whether an individual gene has higher standard deviation than average:
```{r}
hypothesis(fit2,"exp(r_geneID__sigma[Bra000002,Intercept] + b_sigma_Intercept) > exp(b_sigma_Intercept) ",class=NULL)
```

For all genes:

First define the tests we want to do
```{r}
tests <- parnames(fit2) %>% 
  grep("r_geneID__sigma", . ,value=TRUE) %>%
  paste("exp(", . ," + b_sigma_Intercept) > exp(b_sigma_Intercept)", sep="")
head(tests)
```

run the test
```{r}
test.results <- hypothesis(fit2,tests,class=NULL)
```

look at the results
```{r}
rownames(test.results$hypothesis) <- rownames(test.results$hypothesis) %>%
  regmatches(. , regexpr("Bra[0-9]*",.)) 
head(test.results,30)
```

## Reality check: plot the data

separate into different facets for the genes with "significantly" higher variance.  Center (but don't scale) to make for easier viewing.
```{r}
voom.small$Star <- test.results$hypothesis[voom.small$geneID,"Star"]
pl <- voom.small %>%
  filter(!is.na(Star)) %>%
  group_by(geneID) %>%
  mutate(centered=expression-mean(expression)) %>%
  ggplot(aes(x=geneID,y=centered,shape=Star,color=Star)) +
  geom_point() +
  facet_wrap(~ Star, scales="free_x")
pl
```


## Next steps

* Incorporate genotype and treatment samples
* Learn more about how the negative binomial should be fit, especially w.r.t cpm.

# Now include multiple conditions...

60 genes, both two genotypes, two conditions
```{r}
voom.small2 <- voom.tidy %>% filter(startsWith(geneID,"Bra0000"))
hist(voom.small2$expression)
dim(voom.small2)
```

fit the interaction model
```{r, message=FALSE}
system.time(fit3 <- brm(bf(expression ~ (gt*trt|geneID), sigma ~ (1|geneID)),
            prior=c(
              set_prior("normal(4,1)","Intercept"),
              set_prior("student_t(3,0,1)","sd")),
            data = voom.small2,
            iter=10000,
            warmup = 2000,
            control=list(adapt_delta=0.95)
))
```

```{r}
fit3
```

fit the additive model
```{r, message=FALSE}
system.time(fit4 <- brm(bf(expression ~ (gt+trt|geneID), sigma ~ (1|geneID)),
            prior=c(
              set_prior("normal(4,1)","Intercept"),
              set_prior("student_t(3,0,1)","sd")),
            data = voom.small2,
            iter=10000,
            warmup = 2000,
            control=list(adapt_delta=0.95)
))
```

```{r}
fit4
```

```{r}
loo(fit3,fit4)
```

Might as well go with the simpler model, not much difference.

test for importance of variable sigma per gene:

```{r}
system.time(fit5 <- brm(bf(expression ~ (gt+trt|geneID)),
            prior=c(
              set_prior("normal(4,1)","Intercept"),
              set_prior("student_t(3,0,1)","sd"),
              set_prior("student_t(3,0,1)","sigma")
              ),
            data = voom.small2,
            iter=10000,
            warmup = 2000,
            control=list(adapt_delta=0.95)
))
```

```{r}
fit5
```

```{r}
loo(fit4,fit5)
```

Fit 4 (with the gene-specific variance) is strongly preferred.

### explore the fitted parameters

Get the estimated sigmas for each gene:

```{r}
str(ranef(fit4)) #we want the second one
# also, to get back on the original scale we need to take the exponential: see vignette("brms_distreg")
sigma.gene2 <- ranef(fit4)[[2]]
hist(sigma.gene2) #how much the log_sigma is being adjusted for each gene
hist(exp(sigma.gene2)) #in sd units
hist(exp(sigma.gene2+fixef(fit4)["sigma_Intercept",])) #actual sigma being used for each gene
```

Get the estimated mean expression for each gene __note that this is only for the "reference" conditions, which in this case is gtIMB211 and trtDP
```{r}
mean.gene2 <- ranef(fit4)[[1]][,"Intercept"] + fixef(fit4)["Intercept",]
```

Is there a relationship between expression level and standard deviation?  We hope that voom took care of this:

```{r}
mean.var2 <- data.frame(mean=mean.gene2,sigma=exp(sigma.gene2+fixef(fit4)["sigma_Intercept",]))
colnames(mean.var2) <- c("mean","sigma")
ggplot(mean.var2,aes(x=mean,y=sigma)) + geom_point() + geom_smooth()
```
Seems reasonable; no strong trend

To test whether an individual gene has higher standard deviation than average:
```{r}
hypothesis(fit4,"exp(r_geneID__sigma[Bra000002,Intercept] + b_sigma_Intercept) > exp(b_sigma_Intercept) ",class=NULL)
```

For all genes:

First define the tests we want to do
```{r}
tests2 <- parnames(fit4) %>% 
  grep("r_geneID__sigma", . ,value=TRUE) %>%
  paste("exp(", . ," + b_sigma_Intercept) > exp(b_sigma_Intercept)", sep="")
head(tests2)
```

run the test
```{r}
test.results2 <- hypothesis(fit4,tests,class=NULL)
```

look at the results
```{r}
rownames(test.results2$hypothesis) <- rownames(test.results2$hypothesis) %>%
  regmatches(. , regexpr("Bra[0-9]*",.)) 
head(test.results2,30)
```

## Reality check: plot the data

separate into different facets for the genes with "significantly" higher variance.  Center (but don't scale) to make for easier viewing.

compare "stars" for the two data sets (note: same genes but many more samples and conditions for the second 2)
```{r,fig.width=12}
voom.small2$Star <- test.results$hypothesis[voom.small2$geneID,"Star"]
voom.small2$Star2 <- test.results2$hypothesis[voom.small2$geneID,"Star"]

voom.small2$Stars <- ifelse(voom.small2$Star == "*" & voom.small2$Star2 == "*", "both", 
                            ifelse(voom.small2$Star == "*", "small data",
                                   ifelse(voom.small2$Star2 == "*", "large data", "neither")))

pl <- voom.small2 %>%
  filter(!is.na(Star)) %>%
  group_by(geneID,gt,trt) %>%
  mutate(centered=expression-mean(expression)) %>%
  ggplot(aes(x=geneID,y=centered,shape=Stars,color=Stars)) +
  geom_point() +
  facet_grid(gt ~ trt)
pl
```

Interesting...there is definitly condition dependence on the variance.  I wonder if that is due to expression level differences between the conditions or something else?

Plot it without centering:

```{r,fig.width=12}
pl <- voom.small2 %>%
  filter(!is.na(Star)) %>%
  ggplot(aes(x=geneID,y=expression,shape=Stars,color=Stars)) +
  geom_point() +
  facet_grid(gt ~ trt)
pl
```


# Code below here not working, do not run




## starting with neg binomial

start with just a single sample type

```{r, eval=FALSE}
imbdp <- data.tidy %>% filter(gt=="IMB211" & trt=="DP" & startsWith(geneID,"Bra0000"))
imbdp
```

```{r,eval=FALSE}
hist(imbdp$cpm)
```

```{r,eval=FALSE}
get_prior(cpm ~ (1|geneID), family="negbinomial", data=imbdp)
```

### Fit standard negative binomial model

Intercept (i.e. expression level) varies as a function of geneID (as a random effect)

```{r, eval=FALSE}
imbdp$cpm.round <- round(imbdp$cpm,0) # maybe I shouldn't be using cpm at all
fit3 <- brm(cpm.round ~ (1 | geneID),family="negbinomial",
            prior=set_prior("normal(5,5)", "Intercept"),
            data=imbdp,
            iter=10000)
```

```{r, eval=FALSE}
fit3 
plot(fit3)
```

```{r,eval=FALSE}
get_prior(bf(cpm.round ~ (1 | geneID),shape ~ (1|geneID)),family="negbinomial",data=imbdp)
```


```{r,eval=FALSE}
fit4 <- brm(bf(cpm.round ~ (1 | geneID),shape ~ (1|geneID)),
            family="negbinomial",
            prior=set_prior("normal(5,5)", "Intercept"),
            data=imbdp,
            iter=2000,
            control = list(max_treedepth = 20))
```

```{r,eval=FALSE}
fit4
plot(fit4)
```


```{r,eval=FALSE}
loo(fit3,fit4)
```


